---
title: "final_project"
author: Student 1 & Student 2
output: pdf_document
date: "2024-12-01"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
public_data <- read.csv('./data/NA_HQ_public_data.csv')
svi_data <- read.csv('./data/SVI2018_US.csv')
us_cities <- read.csv('./data/uscities.csv')
library(pander)
```

# 0. Contribution Statement

## Student 1

Student 1 mostly worked on questions ...

## Student 2

Student 2 mostly worked on questions ...

\pagenumbering{arabic}

# Introduction

Cities provide a glimpse into the local population of an area and are reflective of the culture and lifestyles in the region in which they are located. However, not all cities are created equal, as different socieoeconomic factors within the population and the surrounding region impact the quality of life of residents.

### Data

Our dataset consists of

### Objective

Our objective is to analyze and document the effects of a 

\pagebreak

# Basic Analysis

## Question 1: SIMPLE REGRESSION

### Methods

**Data Cleaning & Preparation**

```{r echo=FALSE}
pander(quantile(svi_data$EP_UNEMP, p = c(0, 0.25, 0.5, 0.75, 1)) )
sum(svi_data$EP_UNEMP == -999)
# We seem to get some -999, which doesn't make sense in the context of estimated unemployment rates. In this column, we have 546 values / 72837 observations encoded as -999. 

pander(quantile(svi_data$EP_PCI, p = c(0, 0.25, 0.5, 0.75, 1)) )
sum(svi_data$EP_PCI == -999)
# The EP_PCI column also has 481 values encoded as -999. 

# Change the -999 to NaNs
svi_data$EP_UNEMP[svi_data$EP_UNEMP == -999] <- NA
svi_data$EP_PCI[svi_data$EP_PCI == -999] <- NA

# Filter out the NaNs and store them in another dataset called "data".
data <- svi_data[complete.cases(svi_data$EP_PCI, svi_data$EP_UNEMP), ]
```

We want to create a regression model for `EP_PCI`, the estimated per capita income, based on the estimated proportion of unemployment, `EP_UNEMP`. We are predicting `EP_PCI` because it describes the *rate* of unemployment and is normalized against population, unlike `E_PCI`, which describes the estimate *count* the unemployed.

We will check the conditions for fitting a simple linear regression: linearity, homoscedasticity, independence, and normality.

1.  **Linearity** - Is there a linear relationship between our explanatory variable, `EP_UNEMP` and our response variable, `EP_PCI`?

```{r echo=FALSE}
plot(data$EP_UNEMP, data$EP_PCI, 
     xlab = 'Estimated Unemployment Rate', 
     ylab = 'Estimated Per Capita Income', 
     main = 'Trend in Estimated Per Capita Income Based on Estimated Unemp. Rate')
```

From our scatterplot, there appears to be a nonlinear negative trend between unemployment rate and estimated per capita income. As the estimated unemployment rate increases, the estimated per capita income exponentially decreases.

2.  **Homoscedasticity**: Would the variance for the estimated unemployment rate be constant if we were to use a linear model?

```{r echo=FALSE}
model <- lm(EP_PCI ~ EP_UNEMP, data = data)
res <- residuals(model)
plot(data$EP_UNEMP, res, 
     xlab = 'Estimated Unemployment Rate', 
     ylab = 'Estimated Per Capita Income', 
     main = 'Trend in Estimated Per Capita Income Based on Estimated Unemp. Rate')
abline(0,0, col = 'red')
```

The residuals do *not* have constant variance and so, the data is not homoscedastic.

3.  **Independence**: Because the residuals exhibit a clear pattern, independence may be violated.

4.  **Normality**: A QQ plot of our residuals shows our data deviates from the y = x line. Its slight curve upward suggests the distribution of our residuals are right-skewed.

```{r echo = FALSE}
qqnorm(res)
qqline(res, col = 'red')
```

### Analysis

```{r echo = FALSE}
# double log scale
data$log_unemp <- log(data$EP_UNEMP + 1e-6)
data$log_pci <- log(data$EP_PCI + 1e-6)

log_model <- lm(log_pci ~ log_unemp, data = data)

# change predictions back to the original scale
predictions <- predict(log_model, newdata = data.frame(log_unemp = data$log_unemp))
predictions <- exp(predictions)

# Plot the scatterplot
plot(data$log_unemp, data$EP_PCI,
     main = 'Estimated Per Capita Income as a Function of Log(Estimated Populated Rate)',
     xlab = 'Log(Estimated Population Rate)',
     ylab = 'Estimated Per Capita Income')

lines(data$log_unemp, predictions, col='red')

# Model evaluation
# evaluating log model
res <- residuals(log_model)
mse <- mean(res ** 2) # 0.201

# evaluating original model for interpretation purposes
mse_original <- mean((data$EP_PCI - predictions)^2)
```

With our log-transformed model, we have an MSE of about 0.201, which represents the averages squared difference between the actual values of the estimated per capita income and the predicted values. This number is pretty low, which indicates our model performed well. However, the MSE of our original model is 301,671,723, which means our model performed very poorly without the log transformation. 

### Conclusion
Based on these findings, a linear model is not the most optimal model for our question, and a transformation is necessary to accurate fit and predict our data. 