---
title: "final_project"
author: Student 1 & Student 2
output: pdf_document
date: "2024-12-01"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
public_data <- read.csv('./data/NA_HQ_public_data.csv')
svi_data <- read.csv('./data/SVI2018_US.csv')
us_cities <- read.csv('./data/uscities.csv')
```

# 0. Contribution Statement

## Student 1

Student 1 mostly worked on questions ...

## Student 2

Student 2 mostly worked on questions ...

\pagenumber{arabic}

# Introduction
[write stuff here]

### Data
[write stuff here]

### Objective
[write stuff here]

\pagebreak

# Basic Analysis

## Question 1: SIMPLE REGRESSION
### Methods
**Data Cleaning & Preparation**
```{r echo=FALSE}
quantile(svi_data$EP_UNEMP, p = c(0, 0.25, 0.5, 0.75, 1)) 
sum(svi_data$EP_UNEMP == -999)
# We seem to get some -999, which doesn't make sense in the context of estimated unemployment rates. In this column, we have 546 values / 72837 observations encoded as -999. 

quantile(svi_data$EP_PCI, p = c(0, 0.25, 0.5, 0.75, 1)) 
sum(svi_data$EP_PCI == -999)
# The EP_PCI column also has 481 values encoded as -999. 

# Change the -999 to NaNs
svi_data$EP_UNEMP[svi_data$EP_UNEMP == -999] <- NA
svi_data$EP_PCI[svi_data$EP_PCI == -999] <- NA

# Filter out the NaNs and store them in another dataset called "data".
data <- svi_data[complete.cases(svi_data$EP_PCI, svi_data$EP_UNEMP), ]
```

We want to create a regression model for ```EP_PCI```, the estimated per capita income, based on the estimated proportion of unemployment, ```EP_UNEMP```. We are predicting ```EP_PCI``` because it describes the _rate_ of unemployment and is normalized against population, unlike ```E_PCI```, which describes the estimate _count_ the unemployed.

We will check the conditions for fitting a simple linear regression: linearity, homoscedasticity, independence, and normality.
1. **Linearity** - Is there a linear relationship between our explanatory variable, ```EP_UNEMP``` and our response variable, ```EP_PCI```?
```{r echo=FALSE}
plot(data$EP_UNEMP, data$EP_PCI, 
     xlab = 'Estimated Unemployment Rate', 
     ylab = 'Estimated Per Capita Income', 
     main = 'Trend in Estimated Per Capita Income Based on Estimated Unemp. Rate')
```
From our scatterplot, there appears to be a nonlinear negative trend between unemployment rate and estimated per capita income. As the estimated unemployment rate increases, the estimated per capita income exponentially decreases. 
2. **Homoscedasticity**: Would the variance for the estimated unemployment rate be constant if we were to use a linear model?
```{r echo=FALSE}
model <- lm(EP_PCI ~ EP_UNEMP, data = data)
res <- residuals(model)
plot(data$EP_UNEMP, res, 
     xlab = 'Estimated Unemployment Rate', 
     ylab = 'Estimated Per Capita Income', 
     main = 'Trend in Estimated Per Capita Income Based on Estimated Unemp. Rate')
abline(0,0, col = 'red')
```
The residuals do _not_ have constant variance and so, the data is not homoscedastic.
3. **Independence**: Because the residuals exhibit a clear pattern, independence may be violated.
4. **Normality**: A QQ plot of our residuals shows our data deviates from the y = x line. Its slight curve upward suggests the distribution of our residuals are right-skewed.
```{r echo = FALSE}
qqnorm(res)
qqline(res, col = 'red')
```

### Analysis

```{r echo = FALSE}
data$log_unemp <- log(data$EP_UNEMP)
log_model <- lm(log_unemp ~ EP_PCI, data = data) # WHERE THE ERRORS START ROLLING IN

predictions <- predict(log_model, newdata = data.frame(EP_UNEMP = data$EP_UNEMP))

# Plot the scatterplot
plot(data$EP_UNEMP, data$EP_PCI,
     main = 'Gamma Ray Intensity as a Function of the Density of Polyethylene Blocks',
     xlab = 'Density of Polyethylene Blocks',
     ylab = 'Gamma Ray Intensity')

lines(data$EP_UNEMP, predictions, col='red')

# Model evaluation
r_squared <- 1 - (rss / tss) # 0.998
mse <- mean(res ** 2) # 35.345
res <- residuals(log_model)
rss <- sum(res ** 2) # residual sum of squares
tss <- sum((data$EP_PCI - mean(data$EP_UNEMP)) ** 2)
r_squared <- 1 - (rss / tss) # 0.998
mse <- mean(res ** 2)
```
